---
title: "Programming Individual Assignment"
output: html_document
date: "2024-12-03"
---

Import Packages

```{r}
library(readr)
library(tidyverse)
library(reshape2)
library(caret)
library(RColorBrewer)
library(stats)
library(cluster)
library(ggplot2)
library(scales)
library(gridExtra)
library(psych)
library(dplyr)
```

Import Dataset

```{r setup, include=FALSE}
# Download the dataset from GitHub
url <- "https://raw.githubusercontent.com/fongbubble/UoB_MGRCM0034_Car_Sales/main/car_sales.csv"
csv_file_path <- tempfile(fileext = ".csv")
download.file(url, destfile = csv_file_path)
print(paste("CSV File Path:", csv_file_path))

# Read the CSV file
df <- read_csv(csv_file_path)
head(df, 5)
```

Dataset information
```{r}
#Explore the dataset (df)
str(df)
```

Checks for NaNs 
```{r}
# Count the number of NaNs in each column
colSums(is.na(df))
```
Checks for Duplicate
```{r}
# Count the number of duplicate rows
sum(duplicated(df))
```

Descriptive Visualisation

Customer Demographic

```{r}
# Define a custom color palette for genders
custom_palette <- c('Male' = 'steelblue', 'Female' = 'salmon')

# Calculate gender distribution
gender_counts <- df %>% count(Gender)

# Create a pie chart to visualise the gender distribution
ggplot(gender_counts, aes(x = "", y = n, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = custom_palette) +
  labs(title = "Gender Distribution", fill = "Gender") +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.text = element_text(size = 14)
  ) +
  geom_text(aes(label = scales::percent(n / sum(n), accuracy = 0.1)), 
            position = position_stack(vjust = 0.5), size = 5)
```

```{r}
#set up a new dataset
car_data <- read_csv(csv_file_path)

# Create the histogram with KDE
plot <- ggplot(car_data, aes(x = `Annual Income`)) +
  geom_histogram(bins = 290, fill = "pink", alpha = 3) +
  geom_density(aes(y = ..count..), color = "steelblue", size = 0.5) +
  labs(x = "Annual Income (10 Millions)", y = "Count", title = "Annual Income Distribution") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = element_text(size = 20, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_text(size = 30),
    axis.title.y = element_text(size = 30),
    plot.title = element_text(size = 30, hjust = 0.5)
  ) +
  scale_x_continuous(name = "Annual Income", limits = c(0, 1000000), breaks = seq(0, 11200000, by = 50000))

# Print the plot
print(plot)

# Save the plot as a PNG file
ggsave("annual_income_distribution.png", plot = plot, width = 10, height = 8, dpi = 300)
```

Car in a market

```{r}

# Load the new dataset 
car_data <- read_csv(csv_file_path)

# Create a custom function to wrap long text labels
wrap_text <- function(text, width) {
  sapply(text, function(x) paste(strwrap(x, width = width), collapse = "\n"))
}

# Plot 1: Distribution of Price with trendline
plot1 <- ggplot(car_data, aes(x = `Price ($)`)) +
  geom_histogram(bins = 50, fill = "#993366", alpha = 0.7) +
  labs(title = "Distribution of Car Price ($)", x = "Price ($)", y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 40),
        axis.title.x = element_text(size = 40),
        axis.title.y = element_text(size = 40),
        axis.text.x = element_text(size = 40),
        axis.text.y = element_text(size = 40))

# Plot 2: Pie chart for Body Style distribution
body_style_counts <- car_data %>% count(`Body Style`)
plot2 <- ggplot(body_style_counts, aes(x = "", y = n, fill = `Body Style`)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("skyblue", "lightblue", "deepskyblue", "dodgerblue", "steelblue")) +
  labs(title = "Distribution of Vehicles by Body Style", fill = "Body Style") +
  theme_void() +
  theme(plot.title = element_text(size = 40),
        legend.title = element_text(size = 40),
        legend.text = element_text(size = 40)) +
  geom_text(aes(label = scales::percent(n / sum(n))), position = position_stack(vjust = 0.5), size = 13)

# Plot 3: Stacked bar chart for Color and Body Style
color_body_pivot <- car_data %>% count(Color, `Body Style`) %>% spread(`Body Style`, n, fill = 0)
color_body_pivot_long <- color_body_pivot %>% gather(`Body Style`, value, -Color)
plot3 <- ggplot(color_body_pivot_long, aes(x = Color, y = value, fill = `Body Style`)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = scales::hue_pal()(length(unique(car_data$`Body Style`)))) +
  labs(title = "Vehicle Colors by Body Style", x = "Color", y = "Number of Vehicles") +
  theme_minimal() +
  theme(plot.title = element_text(size = 40),
           axis.title.x = element_text(size = 40),
        axis.title.y = element_text(size = 40),
        axis.text.x = element_text(angle = 50, hjust = 1, size = 30),  # Increased size
        axis.text.y = element_text(size = 40),  # Increased size
        legend.title = element_text(size = 40),
        legend.text = element_text(size = 40))

# Plot 4: Pie Chart for Transmission (Auto vs Manual)
transmission_counts <- car_data %>% count(Transmission)
plot4 <- ggplot(transmission_counts, aes(x = "", y = n, fill = Transmission)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("skyblue", "steelblue")) +
  labs(title = "Transmission Proportion: Auto vs Manual", fill = "Transmission") +
  theme_void() +
  theme(plot.title = element_text(size = 40),
        legend.title = element_text(size = 40),
        legend.text = element_text(size = 40)) +
  geom_text(aes(label = scales::percent(n / sum(n))), position = position_stack(vjust = 0.5), size = 13)

# Arrange all plots together
grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)


```

Descriptive analysis

```{r}
# Function to calculate mode
calculate_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Generate descriptive statistics for the selected features
descriptive_stats <- df %>%
  summarise(
    Variable = c("Annual Income", "Price ($)", "Unique Brands", "Unique Models"),
    Mean = c(mean(`Annual Income`, na.rm = TRUE), mean(`Price ($)`, na.rm = TRUE), n_distinct(Company), n_distinct(Model)),
    Median = c(median(`Annual Income`, na.rm = TRUE), median(`Price ($)`, na.rm = TRUE), NA, NA),
    Mode = c(calculate_mode(`Annual Income`), calculate_mode(`Price ($)`), NA, NA),
    SD = c(sd(`Annual Income`, na.rm = TRUE), sd(`Price ($)`, na.rm = TRUE), NA, NA),
    Min = c(min(`Annual Income`, na.rm = TRUE), min(`Price ($)`, na.rm = TRUE), NA, NA),
    Max = c(max(`Annual Income`, na.rm = TRUE), max(`Price ($)`, na.rm = TRUE), NA, NA)
  )

print(descriptive_stats)

```
```{r}
# Count the number of unique brands
num_unique_brands <- n_distinct(df$Company)
print(paste("Number of unique brands:", num_unique_brands))

# Count the number of unique models
num_unique_models <- n_distinct(df$Model)
print(paste("Number of unique models:", num_unique_models))
```

Correlation Heat Map

```{r}
# Copy the data frame
encoded_df <- df

# Encode categorical variables
for (column in colnames(encoded_df)) {
  if (is.factor(encoded_df[[column]]) || is.character(encoded_df[[column]])) {
    encoded_df[[column]] <- as.numeric(factor(encoded_df[[column]]))
  }
}

# Calculate the correlation matrix
corr_matrix <- cor(encoded_df, use = "complete.obs")

# Plot the correlation heatmap
heatmap_data <- melt(corr_matrix)
ggplot(data = heatmap_data, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "Spectral", direction = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap for All Encoded Variables", fill = "Correlation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3)
```

Boxplot

```{r}
# Create the boxplot
ggplot(df, aes(x = Dealer_Region, y = `Price ($)`)) +
  geom_boxplot(aes(fill = Dealer_Region)) +
  scale_fill_brewer(palette = "Spectral") +
  labs(title = "Effect of Dealer Region on Car Price",
       x = "Dealer Region",
       y = "Price ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Perform linear regression
model <- lm(`Price ($)` ~ Dealer_Region, data = df)

# Print the summary of the model
summary(model)
```

Label Encoder

```{r}
# Columns to encode
columns_to_encode <- c('Gender', 'Transmission', 'Body Style', 'Company', 'Model')

# Encode columns
for (col in columns_to_encode) {
  if (col %in% colnames(df) && !all(is.na(df[[col]]))) {
    df[[paste0(gsub(" ", "_", col), '_Encoded')]] <- as.numeric(factor(df[[col]]))
  } else {
    warning(paste("Column", col, "is either missing or contains only NA values."))
  }
}

# Verify the encoded columns
head(df)
```

StandardScaler

```{r}
# Add Gender, Transmission, and Body Style to the feature matrix
X1 <- df %>%
  select(`Annual Income`, `Price ($)`, Company_Encoded, Model_Encoded)

# Scale the features
scaler <- preProcess(X1, method = c("center", "scale"))
X1_scaled <- predict(scaler, X1)

# View the first few rows of the scaled feature matrix
head(X1_scaled)
```

Scatter plot

```{r}
# Create the scatter plot
ggplot(df, aes(x = `Annual Income`, y = `Price ($)`)) +
  geom_point(color = 'steelblue') +
  labs(title = 'Scatter plot of Annual Income vs. Price ($)',
       x = 'Annual Income',
       y = 'Price ($)') +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma)
```



```{r}
#reverse values in X1_pca
X1_pca <- as.data.frame(X1_pca)

X1_pca$PC2 <- -X1_pca$PC2
print(X1_pca)

```

Performing hyperparameter tuning for K-Means using caret package

```{r}

# Define a function to perform K-Means clustering and return the total within-cluster sum of squares
kmeans_wss <- function(k, data, nstart = 25, iter.max = 300, algorithm = "Hartigan-Wong") {
  kmeans(data, centers = k, nstart = nstart, iter.max = iter.max, algorithm = algorithm)$tot.withinss
}

# Define parameter grid for hyperparameter tuning
param_grid <- expand.grid(
  centers = c(3, 4, 5, 6, 7, 8, 9),
  nstart = c(25, 50),
  iter.max = c(300),
  algorithm = c("Hartigan-Wong", "Lloyd", "Forgy", "MacQueen")
)

# Perform grid search to find the best parameters for K-Means
results <- apply(param_grid, 1, function(params) {
  k <- as.numeric(params["centers"])
  nstart <- as.numeric(params["nstart"])
  iter.max <- as.numeric(params["iter.max"])
  algorithm <- as.character(params["algorithm"])
  
  wss <- kmeans_wss(k, X1_pca, nstart = nstart, iter.max = iter.max, algorithm = algorithm)
  c(k, nstart, iter.max, algorithm, wss)
})

# Convert results to a data frame
results_df <- as.data.frame(t(results))
colnames(results_df) <- c("centers", "nstart", "iter.max", "algorithm", "wss")

# Find the best parameters (minimum within-cluster sum of squares)
best_params <- results_df[which.min(results_df$wss), ]
print("Best Parameters:")
print(best_params)

# Use the best model identified by grid search
best_algorithm <- kmeans(X1_pca, centers = as.numeric(best_params["centers"]), nstart = as.numeric(best_params["nstart"]), iter.max = as.numeric(best_params["iter.max"]), algorithm = as.character(best_params["algorithm"]))
```

Feature Contribution

```{r}
# Assuming components is a matrix with feature contributions
# and features is a vector of feature names
components_df <- as.data.frame(t(components))
colnames(components_df) <- features
components_df$Principal_Component <- rownames(components_df)

# Convert the data frame to long format for ggplot2
components_long <- melt(components_df, id.vars = "Principal_Component")
colnames(components_long) <- c("Principal_Component", "Feature", "Contribution")

# Define spectral colors
spectral_colors <- brewer.pal(n = length(features), name = "Spectral")

# Create the bar plot
ggplot(components_long, aes(x = Principal_Component, y = Contribution, fill = Feature)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = spectral_colors) +
  labs(title = "Feature Contribution to Each Principal Component",
       x = "Principal Components",
       y = "Contribution") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.position = "top",
        legend.box.background = element_rect(color = "black", fill = "white", size = 0.5)) +
  theme(plot.title = element_text(hjust = 0.5))
```

GridSearch to tune hyper parameter

```{r}
# Define parameter grid for hyper parameter tuning
param_grid <- expand.grid(
  n_clusters = c(3, 4, 5, 6, 7, 8, 9),
  init = c('k-means++', 'random'),
  max_iter = c(300)
)

# Function to perform K-Means clustering and return total within-cluster sum of squares
kmeans_clustering <- function(data, centers, nstart, iter.max) {
  kmeans(data, centers = centers, nstart = nstart, iter.max = iter.max)$tot.withinss
}

# Perform grid search manually
best_score <- Inf
best_params <- NULL

for (i in 1:nrow(param_grid)) {
  params <- param_grid[i, ]
  score <- kmeans_clustering(
    data = X1_pca,
    centers = params$n_clusters,
    nstart = ifelse(params$init == 'k-means++', 10, 1),
    iter.max = params$max_iter
  )
  
  if (score < best_score) {
    best_score <- score
    best_params <- params
  }
}

print("Best Parameters:")
print(best_params)
```

PCA with K-means clustering

```{r}
# Fit the best K-Means algorithm and assign cluster labels to DataFrame
best_k <- best_params$n_clusters
best_init <- best_params$init
best_max_iter <- best_params$max_iter

set.seed(24150)
kmeans_result <- kmeans(
  X1_pca, 
  centers = best_k, 
  nstart = ifelse(best_init == 'k-means++', 10, 1), 
  iter.max = best_max_iter
)

# Ensure the Cluster column is correctly assigned
df$Cluster <- kmeans_result$cluster

# Define descriptive names for each cluster
custom_cluster_names <- c(
  "High-Income, Luxury Seekers", 
  "Mid-Income, Practical Buyers", 
  "High-Income, Brand-Focused Buyers",
  "Mid-Income, Budget-Conscious Buyers",
  "Low-Income, Budget-Focused Buyers",
  "High-Income, Cost-Conscious Buyers",
  "Mid-Income, Brand-Focused Buyers",
  "Low-Income, Value-Driven Buyers",
  "Low-Income, Practical Buyers"
)

# Create a mapping from cluster numbers to descriptive names
cluster_mapping <- setNames(custom_cluster_names, 1:best_k)
df$ClusterName <- cluster_mapping[df$Cluster]

# Generate car brand and model recommendations for each cluster
cluster_recommendations <- list()
for (cluster in 1:best_k) {
  cluster_data <- df[df$Cluster == cluster, ]
  recommended_brands <- head(sort(table(cluster_data$Company), decreasing = TRUE), 3)
  brand_model_recommendations <- list()
  for (brand in names(recommended_brands)) {
    top_models <- head(sort(table(cluster_data[cluster_data$Company == brand, ]$Model), decreasing = TRUE), 1)
    brand_model_recommendations[[brand]] <- names(top_models)
  }
  cluster_recommendations[[cluster]] <- brand_model_recommendations
}

# Print the car brand and model recommendations for each cluster
cat("Car Brand and Model Recommendations for Each Cluster:\n")
for (cluster in 1:best_k) {
  cluster_name <- custom_cluster_names[cluster]
  cat(paste0(cluster_name, ":\n"))
  for (brand in names(cluster_recommendations[[cluster]])) {
    models <- paste(cluster_recommendations[[cluster]][[brand]], collapse = ", ")
    cat(paste0("  ", brand, ": ", models, "\n"))
  }
}

# Create mesh grid for visualizing K-Means clustering boundaries
x_min <- min(X1_pca[, 1]) - 1
x_max <- max(X1_pca[, 1]) + 1
y_min <- min(X1_pca[, 2]) - 1
y_max <- max(X1_pca[, 2]) + 1
xx <- seq(x_min, x_max, by = 0.05)
yy <- seq(y_min, y_max, by = 0.05)
grid <- expand.grid(Var1 = xx, Var2 = yy)

# Manually assign cluster labels to each point in the mesh grid
assign_cluster <- function(point, centroids) {
  distances <- apply(centroids, 1, function(centroid) sum((point - centroid)^2))
  return(which.min(distances))
}

centroids1 <- kmeans_result$centers
grid$Cluster <- apply(grid, 1, assign_cluster, centroids = centroids1)
grid$ClusterName <- cluster_mapping[grid$Cluster]

# Define the order of clusters for the legend
cluster_order <- c(
  "High-Income, Luxury Seekers",
  "High-Income, Brand-Focused Buyers",
  "High-Income, Cost-Conscious Buyers",
  "Mid-Income, Brand-Focused Buyers",
  "Mid-Income, Practical Buyers",
  "Mid-Income, Budget-Conscious Buyers",
  "Low-Income, Value-Driven Buyers",
  "Low-Income, Practical Buyers",
  "Low-Income, Budget-Focused Buyers"
)

# Plot PCA scatter plot with K-Means clustering
library(ggplot2)
library(RColorBrewer)

ggplot() +
  geom_tile(data = grid, aes(x = Var1, y = Var2, fill = factor(ClusterName, levels = cluster_order)), alpha = 0.3) +
  geom_point(data = as.data.frame(X1_pca), aes(x = PC1, y = PC2, color = factor(df$ClusterName, levels = cluster_order)), alpha = 0.8) +
  geom_point(data = as.data.frame(centroids1), aes(x = PC1, y = PC2), color = 'red', size = 3, shape = 8) +
  scale_fill_manual(values = brewer.pal(n = best_k, name = "Spectral"), breaks = cluster_order) +
  scale_color_manual(values = brewer.pal(n = best_k, name = "Spectral"), breaks = cluster_order) +
  labs(title = 'PCA Scatter Plot with K-Means Clustering',
       x = 'PC1: Car Preference & Affordability Component',
       y = 'PC2: Income Component',
       fill = 'Cluster Labels',
       color = 'Cluster Labels') +
  theme_minimal() +
  theme(legend.position = 'right')

```

In this version, I've explicitly named the columns `Var1` and `Var2` in the `grid` data frame. This should resolve the error and allow `ggplot` to correctly interpret the aesthetics. Let me know if this works! \`\`\`
